---
title: "data_analysis"
author: "Mira Beinart"
date: "2025-04-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)  
library(knitr)      
library(lubridate)
library(janitor)
```

```{r}
detail_sod <- read_csv("data/detail_sod.csv")
```
#fixing dates
```{r}
detail_sod <- detail_sod %>%
  mutate(TRANSACTION_DATE=dmy(TRANSACTION_DATE))
```
```{r}
 detail_sod <- detail_sod %>%
  mutate(PERFORM_START_DT=dmy(PERFORM_START_DT))
```
```{r}
detail_sod <- detail_sod %>%
  mutate(PERFORM_END_DT=dmy(PERFORM_END_DT))
```
```{r}
write_csv(detail_sod, "data/detail_sod_clean.csv")
```

```{r}
detail_sod_clean <- read_csv("detail_sod_clean.csv")

glimpse(detail_sod_clean)
```
information:
- columns: 18
- rows: 129,843
- limitations: the data doesn't include any party affiliations, which would be helpful for analysis down the line (eg. looking at spending trends between parties, etc). also, the data only includes expenditures from the second half of 2024, so we will probably need to find more data to analyze trends over a longer period of time.
**the information is the same in the summary vs the detail dataframe, the difference is the size of the data - the summary dataframe has 5 columns with 6,676 rows

```{r}
summary_sod <- read_csv("data/summary_sod.csv")

glimpse(summary_sod)
```


