---
title: "Reverse Engineering Project"
author: "Mira, Laura ,Katelynn "
date: "March 15, 2025"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story, [Millions in out-of-state donations help fuel high-profile Maryland Democratic governor candidates](https://cnsmaryland.org/2022/03/03/millions-in-out-of-state-donations-help-fuel-high-profile-maryland-democratic-governor-candidates/)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
options(scipen=999)
library(tidyverse)
library(lubridate)
library(janitor)
library(dplyr) 
library(readxl)
```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

**DATA QUESTIONS FOR RYAN** Hi, we were unable to find the contribution data for the Rush Baker campaign through the website, so instead of loading all seven candidates we only have six. 
I'm not sure the data for all of them is even correct because some were like 'friends of candidate' and some were just normal 'candidate for governor/maryland', some even had both. Should we include both if applicable and what should we do about the Baker data? -- On second look unless we want to recreate the big yucky graph, theres no need to include schulz gansler or baker, so is it even necesary?

```{r}
# Load required data

moore <- read_csv("data/wesmoore.csv")

gansler <- read_csv("data/douggansler.csv")

king <- read_csv("data/johnking.csv")

schulz <- read_csv("data/kellyschulz.csv")

perez <- read_csv("data/tomperez.csv")

franchot <- read_csv("data/peterfranchot.csv")

# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")

# Clean required data and prepare for analysis if needed. 

```

## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

* **Sentence text**: Moore, a U.S. Army veteran and bestselling author, built a campaign war chest that included over $2.3 million in out-of-state contributions. 
* **Analysis summary**: We were able to confirm the finding. Out of Wes Moore's 4,127,418 dollars in campaign funding, 2,387,822.72 dollars came from out-of-state donations. Less than 2 million was from Maryland donations, which the story does mention: "Moore raised over $1.7 million in Maryland, the highest total of any candidate," so we replicated that finding too.

```{r}
#cleaning column names
cleaned_moore <- moore |>
clean_names()
```


```{r}
#checking total contrib amount
cleaned_moore |>
group_by(receiving_committee)|>
  summarise(
    total_contrib_amount = sum(contribution_amount)
  )
```


```{r}
#mutating state out of contributor address data

states_moore <- cleaned_moore |>
    mutate(contributor_state = str_extract(contributor_address, "\\b[A-Z]{2}(?=\\s+\\d{5}\\b)"))|>
  
# The mutate code here is from ChatGPT. I entered the data and asked for "a new column that outputs the state within the contributor address column using R tidyverse"

# Here I sorted the state by top contributions per state
group_by(contributor_state)|>
  summarise(
    state_contribs = sum(contribution_amount)
  )|>
  arrange(desc(state_contribs))


print(states_moore)
```

```{r}
 ### Comparing in-state contribution amounts to out-of-state/overseas contribution amounts
  state_contribs_moore <- states_moore|>
  
    mutate(
      contrib_location = case_when(
    is.na(contributor_state) ~ "Overseas", #Code here is from from ChatGPT. I asked how to use 'is.na' or something similar in a case_when mutate function
    contributor_state!="MD" ~ "Out of State",
      contributor_state== "MD" ~ "Maryland"
  ))|>
    
group_by(contrib_location)|>
  summarise(
   total_contribution = sum(state_contribs)
  )|>
  adorn_totals()

print(state_contribs_moore)
```


### Sentence 2

* **Sentence text**: John B. King, former secretary of Education under President Barack Obama, raised $2.2 million in campaign funding last year. Only 11%, just under $240,000, came from Maryland donors. King raised over $534,000 in New York alone.


* **Analysis summary**: Yes, we were able to replicate this finding. King raised exactly 2,199,015 dollars for his campaign. 534478.44 of that came from New York state. Only 238,527 dollars came from Maryland. That is about 11% of his total.

```{r}
#cleaning column names
cleaned_king <- king |>
clean_names()
```


```{r}
#checking total contrib amount
cleaned_king |>
group_by(receiving_committee)|>
  summarise(
    total_contrib_amount = sum(contribution_amount)
  )
```


```{r}
#mutating state out of contributor address data
states_king <- cleaned_king |>
    mutate(contributor_state = str_extract(contributor_address, "\\b[A-Z]{2}(?=\\s+\\d{5}\\b)"))|>

# The mutate code here is from ChatGPT: I entered the data and asked for "a new column that outputs the state within the contributor address column using R tidyverse"

group_by(contributor_state)|>
  summarise(
    state_contribs = sum(contribution_amount)
  )|>
  arrange(desc(state_contribs))


print(states_king)
```

```{r}
 ### comparing in-state contribution amount to out-of-state with raw numbers and percentages
  state_contribs_king <- states_king|>
  
    mutate(
      contrib_location = case_when(
    is.na(contributor_state) ~ "Overseas", #This code is from ChatGPT. I asked how to use is.na or something similar in a case_when mutate function
    contributor_state!="MD" ~ "Out of State",
      contributor_state== "MD" ~ "Maryland"
  ))|>
    
group_by(contrib_location)|>
  summarise(
   total_contribution = sum(state_contribs)
  )|>
   mutate(
    percent_of_total = total_contribution / sum(total_contribution) * 100
  ) |>
  adorn_totals()

print(state_contribs_king)


```


### Sentence 3

* **Sentence text**: Tom Perez, a former leader of the Democratic National Committee, raised nearly $1.7 million in donations from outside Maryland.
* **Analysis summary**: Yes, we were about to replicate this finding. 1,695,893.20 dollars of Perez's campaign contributions came from out of state. 


```{r}
###cleaning column names
cleaned_perez <- perez |>
clean_names()
```

```{r}
#checking total contribution amount
cleaned_perez |>
group_by(receiving_committee)|>
  summarise(
    total_contrib_amount = sum(contribution_amount)
  )
```

```{r}
#mutating state out of contributor address data

states_perez <- cleaned_perez |>
    mutate(contributor_state = str_extract(contributor_address, "\\b[A-Z]{2}(?=\\s+\\d{5}\\b)"))|>

### This mutate code is from ChatGPT. I entered the data and asked for "a new column that outputs the state within the contributor address column using R tidyverse."

group_by(contributor_state)|>
  summarise(
    state_contribs = sum(contribution_amount)
  )|>
  arrange(desc(state_contribs))

print(states_perez)
```

```{r}
 ### comparing in state contribution amount to out-of-state
  state_contribs_perez <- states_perez|>
  
    mutate(
      contrib_location = case_when(
    is.na(contributor_state) ~ "Overseas", #code from chatgpt-- how to use is.na or something similar in a case_when mutate function
    contributor_state!="MD" ~ "Out of State",
      contributor_state== "MD" ~ "Maryland"
  ))|>
    
group_by(contrib_location)|>
  summarise(
   total_contribution = sum(state_contribs)
  )|>
  adorn_totals()

print(state_contribs_perez)
```


### Sentence 4

* **Sentence text**: King’s campaign, which raised just under 11% of its money from Maryland, said it received donations from every county in the state, and said the support it received from other states was indicative of voters looking for a candidate with leadership on education issues.

* **Analysis summary**: We were somewhat able to verify this analysis. After importing a dataset with Maryland zipcodes by county, we wrote code to extract the zipcodes from each contribution and assign it to a county. The result is a dataset with 23 rows (but there are 24 counties in Maryland, including Baltimore City), with one row with a county called NA. St. Mary's and Somerset counties are missing from this list. 

```{r}
# cleaning column names
cleaned_king <- king |>
clean_names()
```

```{r}
# checking total contribution amount
cleaned_king|>
group_by(receiving_committee)|>
  summarise(
    total_contrib_amount = sum(contribution_amount)
  )
```

```{r}
# importing zipcode data to compare addresses to verify that he received contributions from each county in MD
zip_county <- read_excel("data/ZIP_COUNTY_122024.xlsx")

# filtering for just MD zipcodes
md_zip_county <- zip_county |> 
  filter(USPS_ZIP_PREF_STATE == "MD")

# make sure state abbreviation exists for filtering
cleaned_king <- cleaned_king |>
  mutate(contributor_state = str_extract(contributor_address, "\\b[A-Z]{2}\\b"))

# county FIPS-to-name lookup table
md_fips_lookup <- tibble::tibble(
  COUNTY = c(
    1, 3, 5, 9, 11, 13, 15, 17, 19, 21, 23, 25, 27, 29,
    31, 33, 35, 37, 39, 41, 43, 45, 47, 510
  ),
  county_name = c(
    "Allegany", "Anne Arundel", "Baltimore City", "Calvert", "Caroline",
    "Carroll", "Cecil", "Charles", "Dorchester", "Frederick", "Garrett",
    "Harford", "Howard", "Kent", "Montgomery", "Prince George's", "Queen Anne's",
    "St. Mary's", "Somerset", "Talbot", "Washington", "Wicomico", "Worcester", "Baltimore"
  )
)

# ensure both sides of the join use the same type (character)
md_donations_by_county <- md_donations_by_county |> 
  mutate(county_fips = substr(COUNTY, 3, 5))  # extract last 3 digits

md_fips_lookup <- md_fips_lookup |> 
  mutate(county_fips = stringr::str_pad(COUNTY, 3, pad = "0"))  # pad with zeros

# now join on county_fips
md_donations_named <- md_donations_by_county |>
  left_join(md_fips_lookup, by = "county_fips") |>
  select(county_name, total_donations, num_contributors) |>
  arrange(desc(total_donations))

print(md_donations_named)


```

```{r}
# cleaning column names
cleaned_franchot <- franchot |>
clean_names()
```

```{r}
# checking total contribution amount
cleaned_franchot |>
group_by(receiving_committee)|>
  summarise(
    total_contrib_amount = sum(contribution_amount)
  )
```

```{r}
# mutating state out of contributor address data
states_franchot <- cleaned_franchot |>
    mutate(contributor_state = str_extract(contributor_address, "\\b[A-Z]{2}(?=\\s+\\d{5}\\b)"))|>
  
# The mutate here is from code from ChatGPT. I entered the data and asked for "a new column that outputs the state within the contributor address column using R tidyverse"

group_by(contributor_state)|>
  summarise(
    state_contribs = sum(contribution_amount)
  )|>
  arrange(desc(state_contribs))

print(states_franchot)
```

```{r}
 ### comparing in state contribution amount to out-of-state
  state_contribs_franchot <- states_franchot|>
  
    mutate(
      contrib_location = case_when(
    is.na(contributor_state) ~ "Overseas", #The code here from chatgpt-- how to use is.na or something similar in a case_when mutate function
    contributor_state!="MD" ~ "Out of State",
      contributor_state== "MD" ~ "Maryland"
  ))|>
    
group_by(contrib_location)|>
  summarise(
   total_contribution = sum(state_contribs)
  )|>
  adorn_totals()
```

```{r}
1303852.5/1663247.4
```


```{r} 
 ### comparing in state contribution amount to out-of-state
  state_contribs_franchot <- states_franchot |>
  
    mutate(
      contrib_location = case_when(
    is.na(contributor_state) ~ "Overseas", #The code here from chatgpt-- how to use is.na or something similar in a case_when mutate function
    contributor_state!="MD" ~ "Out of State",
      contributor_state== "MD" ~ "Maryland"
  ))|>
    
group_by(contrib_location)|>
  summarise(
   total_contribution = sum(state_contribs)
  )|>
  adorn_totals()

```

```{r}
1303852.5/1663247.4
```

### Sentence 5

* **Sentence text**: An analysis of campaign data by Capital News Service found other candidates in the Democratic primary similarly drew from the well of money from other states. Their campaigns were fattened with donations from Democratic strongholds like New York, California and Chicago, campaign finance records show.

* **Analysis summary**: We were able to recreate this finding. New York, California and Illinois (Chicago) are some of the top contributors for candidates like Moore, Perez and King.


```{r}
colnames(states_king)
colnames(states_moore)
colnames(states_perez)

print(states_king)
print(states_moore)
print(states_perez)
```


-30-