---
title: "lab_01"
author: "Ryan Little"
date: "2025-02-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your reference notebook!

## Load libraries and establish settings

**Task** Import the Tidyverse. Write the code needed to load the library. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
library("tidyverse")
```

## Load Data

In this lab, we will be recreating the first data story I ever did as a data journalism student at Merrill while working in the Howard Center for Investigative Journalism. First, we need to import the court charges data from Okaloosa County, Florida.

**Task** Using the example from the pre_lab, write code to load the CSV file in the data file and assign it to a good variable name. This data is only charges for homeless people.

```{r}
homeless_people <- read.csv("data/okaloosa-court-charges.csv")
```

## Examine the data

**Task** Run the codeblock below to use glimpse to get a sense of the column names, types and sample data.

**Q1** How many rows and columns does this dataframe have?

**A1** 15,386 rows, 56 columns

```{r}
glimpse(homeless_people)
```

## Answer questions

**Q2** Which charges are homeless people most often charged with?

**A2** narcotic equip possess or use

```{r}
homeless_people |>
  group_by(court_statute) |>
  summarise(
    count_charge = n()
  ) |>
arrange(desc(count_charge))
```

**Q3** What charges against homeless people have been charged the least?

**A3** adult permits under 16 yoa to possess weapon, battery against a pregnant woman?!, aggravated stalking, etc.

```{r}
homeless_people |>
  group_by(court_statute) |>
  summarise(
    count_charge = n()
  ) |>
arrange(count_charge)
```

**Q4** Why do you think there are so many charges identified in **A3** that are tied for the least common? Hint: You can't answer this question with code.

**A4** I think because there are a lot of different charges, a lot of them that mean the same thing, but with specific terms that make them ever so slightly different, which sorts them into their own category in the data. Also, the data is all entered by humans, so there's a chance that people use different acronyms and/or make spelling errors.

**Q5** Which prosecutor has prosecuted the most charges against homeless people? Is the top prosecutor related to any famous Floridians?

**A5** Ginger Bowden Madden has prosecuted the most charges against homeless people, at 972. She is the daughter of Bobby Bowden, FSU football coach for over 30 years.

```{r}
homeless_people |>
  group_by(prosecutor) |>
  summarise(
    count_prosecutor = n()
  ) |>
  arrange(desc(count_prosecutor))
```


**Q6** Which judge has overseen the most cases involving homeless people? Hint: Cases and charges are not the same thing.

**A6** Patricia S. Grinsted!

```{r}
homeless_people |>
  group_by(judge) |>
  summarise(
    count_judge = n_distinct(case_number)
  ) |>
arrange(desc(count_judge))
```

**Q7** What prosecutor has prosecuted the most cases against a single homeless person?

**A7** Cassie Reed, but we don't know the prosecutors for the top 5 defendants with the most charges. 

```{r}
homeless_people |>
  group_by(prosecutor, defendant) |>
  summarise(
    count_defendant = n()
  ) |>
arrange(desc(count_defendant))
```

**Q8** We need to create a dataframe that will help us when we are reporting on the ground. Building off our work so far, we need a dataframe that includes the name of every homeless person who was charged with a crime and some important counts about their interactions with the criminal justice system. Create a dataframe that includes the name of each homeless person, the number of charges they have faced, the number of cases they have been involved in, how many dollars they have been assessed in fines, the average fine imposed on them and the median fine imposed upon them. Then sort the data frame so the person with the most charges is at the top. 

**A8** I used ChatGPT to tell me how to create a new dataframe -- %>% and select

```{r}
reporting_homeless_people <- homeless_people%>%
  select(defendant,court_statute, total_assessed, case_number)
```

```{r}
reporting_homeless_people |>
  group_by(defendant, court_statute) |>
    summarise(
      count_defendant = n_distinct(case_number),
      count_court_statute = n(),
      total_fines = sum(total_assessed),
      mean_fines = mean(total_assessed),
      median_fines = median(total_assessed)
    ) |>
arrange(desc(count_defendant))

```

**Q9** Please describe a short reporting plan for how you would use the data you have generated in **A8** to report on the criminal justice system in Okaloosa County, Florida. Who are the people we should talk to? What more should we do with our data analysis to be certain we know what the story is?

**A9** We should talk to Kenneth Ivan Shultz III, Kelliann Marie Clark, and Michael Mario Wilson, as they are the top three prosecuted homeless people in Okaloosa County, FL. We should also look at the geotag data to see where Kenneth and Michael were arrested for trespassing each time to see if there is a specific location where they are trespassing more regularly. We should also look at who is charging them -- if we find a specific officer that is arresting the majority of homeless people, we should find them and interview them as well. I also think we should interview Ginger Bowden, the most prominent prosecutor in this dataset, as well as Patricia Grinsted, the judge who oversaw the most cases. 
