---
title: "lab_10"
author: "Derek Willis"
date: "2024-11-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(tigris_use_cache = TRUE)
install.packages("mapview")
```

## You will need

-   the tidyverse, janitor and tigris libraries
-   you might choose to use the tidycensus library for Q3

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load libraries
library(tidyverse)
library(janitor)
library(tigris)
library(sf)
library(ggplot2)
library(mapview)

```

Let's explore the election results from last night and make some maps!

## Questions

**Q1.** Make a county-level map of Maryland's results for U.S. Senate using `md_senate_county_24.csv` in the data folder, calculating the difference between Angela Alsobrooks's percentage of the total votes and Larry Hogan's percentage of the total votes and displaying that percentage difference on the map. What is the story here, and what is the county with the most interesting results?

**A1.** The majority of Maryland counties favored Hogan, which surprised me. I think the biggest story is Baltimore County where the percentage of votes for Hogan and Alsobrooks is nearly equal, making the county almost white in the graph (ever so slightly light blue). I would go to Baltimore County and find people on both ends of the political spectrum and see why they feel that way, and then probably analyze some more data. Story is brewing. 

```{r}
#load data

md_senate <- read_csv("data/md_senate_county_24.csv") |>
  clean_names() |>
  mutate(total = alsobrooks + hogan + scott, 
         alsobrooks_pct = alsobrooks / total * 100,
         hogan_pct = hogan / total * 100,
         difference = alsobrooks_pct - hogan_pct)
```
```{r}
#getting county data
counties <- counties()
#filtering for just Maryland counties 
md_counties <- counties |>
  filter(STATEFP == "24")
#fix the issue of Baltimore City and County both being listed as just "Baltimore" and the difference in St. Mary's. always that damn county.
md_counties <- md_counties |>
  mutate(NAME = case_when(
    NAME == "Baltimore" & NAMELSAD == "Baltimore city" ~ "Baltimore City",
    NAME == "Baltimore" & NAMELSAD == "Baltimore County" ~ "Baltimore County",
    NAME == "St. Mary's" & NAMELSAD == "St. Mary's County" ~ "Saint Mary's",
    TRUE ~ NAME
  ))


```
```{r}
#join the data
md_map_data <- md_counties |>
  left_join(md_senate, by = c("NAME" = "county"))


ggplot(md_map_data) +
  geom_sf(aes(fill = difference), color = "white") +
  scale_fill_gradient2(
    low = "#d73027", mid = "white", high = "#4575b4",  # red-white-blue scale
    midpoint = 0,
  ) +
  labs(
    title = "Maryland Senate Results by County",
    subtitle = "A red county indicates a higher percentage of votes for Larry Hogan (R), \nand a blue county indicates a higher percentage of votes for Angela Alsobrooks (D).",
    caption = "Source: Maryland Board of Elections"
  ) +
  theme_minimal() +
   theme(
    axis.text = element_blank(),       # remove lat/lon numbers
    axis.ticks = element_blank(),      # remove tick marks
    panel.grid = element_blank()       # remove grid lines #thanks chat for this one!!!!
  )
```
```{r}
#me just fuckin around with mapview 
mapview(md_map_data, zcol = "difference", legend = TRUE) 
```


**Q2.** Make a county-level map showing the difference between Donald Trump's county-level performance this year and Larry Hogan's, using percentages to compare the two as you did in Q1. You can use the dataframe you initially loaded for Q1, and you'll need to load the 2024 presidential results in the data folder and join them to it before proceeding. Are there any counties where Trump got a higher percentage than Hogan? How would you describe the map showing the Trump-Hogan difference?

Also answer this: is a map the best way to present this data? What else could you make that might convey more information?

**A2.** Allegany County is the only county where Trump got a higher percentage of votes than Hogan. I highlighted this by outlining the county in a darker line to draw attention, and including that information in the subtitle/description part of the map. I think a bar chart/graph could be more useful for this data because it would give raw numbers, showing the difference in votes for Trump and Hogan in each county.

```{r}
#load the new data
md_pres <- read_csv("data/md_pres_county_24.csv") |>
  clean_names() |>
  mutate(total = harris + trump + oliver + stein + kennedy,
         trump_pct = trump / total * 100)

#make new dataset form Q1 data using just the information i need
senate_for_pres <- md_senate |>
  select(county, hogan, total) |>
  mutate(real_hogan = hogan / total * 100)

#join the data
pres_map <- md_pres |>
  left_join(senate_for_pres, by = c("county" = "county")) |>
  mutate(trumpy = real_hogan - trump_pct)

#join with county data
mappy_map_pres <- md_counties |>
  left_join(pres_map, by = c("NAME" = "county"))

#make a new dataset for Allegany outline 
allegany_outline <- mappy_map_pres |> 
  filter(NAME == "Allegany")

#make the mf map
ggplot(mappy_map_pres) +
  geom_sf(aes(fill = trumpy), color = "white") +
   geom_sf(data = allegany_outline, fill = NA, color = "black", size = 2.5) +
  scale_fill_gradient(
    low = "#fee5d9",    # light red
    high = "#a50f15",   # dark red
    name = "difference"
  ) +
  labs(
    title = "County-Level Difference: Trump vs. Hogan (2024)",
    subtitle = "Allegany County was the only county that saw a higher percentage of votes for Trump than Hogan.",
    caption = "Source: Maryland Board of Elections"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),       # remove lat/lon numbers
    axis.ticks = element_blank(),      # remove tick marks
    panel.grid = element_blank()       # remove grid lines #thanks chat for this one!!!!
  )

```



**Q3** Make another map showing the difference between Larry Hogan's county-level performance this year and from his governor race in 2018, using percentages to compare the two as you did in Q2. You can use the dataframe you initially loaded for Q1, and you'll need to load the 2018 governor results in the data folder and join them to it before proceeding. Are there any counties where Hogan did better this year? How would you describe the map showing the difference?

**A3** There are no counties where Hogan did better in 2024 than in 2018. The map shows that Hogan's support has decreased in every county, with the most significant drop in Baltimore City. This is interesting considering the nation-wide shift towards Republican candidates that happened this past year.

```{r}
#load 2018 data

gov_2018 <- read_csv("data/md_gov_county_18.csv") |>
  clean_names() |>
  mutate(total = hogan + jealous + quinn + schlakman + write_ins,
         hogan_pct18 = hogan / total * 100)

#the 2018 data had a random weird space in Saint Mary's like literally what I was talking about in class???? wtf????? thank you chat for helping me fix this.
gov_2018 |> filter(str_detect(county, "Mary"))
senate_for_pres |> filter(str_detect(county, "Mary"))

charToRaw(gov_2018$county[which(str_detect(gov_2018$county, "Mary"))])
charToRaw(senate_for_pres$county[which(str_detect(senate_for_pres$county, "Mary"))])

library(stringr)

clean_county <- function(x) {
  x |>
    str_replace_all("\u00A0", " ") |>  # replace non-breaking space with real space
    str_replace_all("’", "'") |>       # smart quote to straight
    str_squish()
}

gov_2018 <- gov_2018 |> mutate(county = clean_county(county))
senate_for_pres <- senate_for_pres |> mutate(county = clean_county(county))

#join the datas
gov_comparison <- gov_2018 |>
  left_join(senate_for_pres, by = c("county" = "county")) |>
  rename(hogan_24 = real_hogan) |>
  rename(hogan_18 = hogan_pct18) |>
  mutate(hogan_difference = hogan_24 - hogan_18)

#make the mf map again
hogan_map_data <- md_counties |>
  left_join(gov_comparison, by = c("NAME" = "county"))

ggplot(hogan_map_data) +
  geom_sf(aes(fill = hogan_difference), color = "white") +
  scale_fill_gradient2(
    low = "#fdae61",
    mid = "white",
    high = "#1a9850",
    name = "Hogan 2024 - 2018 (%)"
  ) +
  labs(
    title = "Hogan saw a higher percentage of votes in every county in 2018 than in 2024",
    subtitle = "The darker the orange fill of the county, the stronger the support Hogan received in 2018 compared to 2024.",
    caption = "Source: Maryland Board of Elections"
  ) +
  coord_sf(datum = NA) +
   theme_minimal() +
  theme(
    axis.text = element_blank(),       # remove lat/lon numbers
    axis.ticks = element_blank(),      # remove tick marks
    panel.grid = element_blank()       # remove grid lines #thanks chat for this one!!!!
  )
```

**Q4.** Choose your own map adventure! In the data folder I've included Maryland county-level results for the abortion rights amendment vote this year, plus the 2020 presidential results by county and 2022 governor's results by county. Using one of those and at least one other contest for comparison, make a county-level map showing what you think is the most interesting aspect of that data, and describe the results in terms of a possible story. You can bring in other data, such as Census information, if you like.

**A4.** I compared the support between Hogan and the abortion question, and the map was really interesting. It shows that Hogan did much better in the counties on the borders of the state, most of which are more rural. However, the abortion question got a higher percentage of votes in more metropolitan areas, like Moco, PG County, and Baltimore City and County.

```{r}
#load abortion data
abortion_24 <- read_csv("data/md_abortion_24.csv") |>
  clean_names() |>
  mutate(total = `for` + against,
         for_pct = `for` / total * 100,
         against_pct = against / total * 100,
         difference = for_pct - against_pct)

#join to senate results from 2024
joined_abortion <- abortion_24 |>
  left_join(md_senate, by = c("county" = "county"))

#join with county data
abortion_map_data <- md_counties |>
  left_join(joined_abortion, by = c("NAME" = "county"))

#adding a column for the difference between for Hogan and for abortion
abortion_map_data <- abortion_map_data |>
  mutate(
    hogan_gap = hogan - `for`,   # if both are raw vote counts
    hogan_gap_pct = hogan_pct - (`for` / (`for` + against) * 100)
  )
#making map
ggplot(abortion_map_data) +
  geom_sf(aes(fill = hogan_gap_pct), color = "white") +
  scale_fill_gradient2(
    low = "#0570b0", mid = "white", high = "#ca0020",
    midpoint = 0,
    name = "Hogan - Abortion Support (%)"
  ) +
  labs(
    title = "Hogan vs. Abortion Support in Maryland",
    subtitle = "Hogan did better compared to the abortion question in more rural counties, \nwhile the abortion question did better in metropolitan, more populated areas.",
    caption = "Source: Maryland Board of Elections"
  ) +
  coord_sf(datum = NA) +
   theme_minimal() +
  theme(
    axis.text = element_blank(),       # remove lat/lon numbers
    axis.ticks = element_blank(),      # remove tick marks
    panel.grid = element_blank()       # remove grid lines #thanks chat for this one!!!!
  )
```

-30-
